Co-evolution of lexical meaning \& pragmatic use

### 2017-04-02 paper/paper-v3.tex
finishing touches first paper draft

\author{%\bf NAME1 and NAME2\\
    ( -- draft \today --- )
}

@@

\begin{abstract}
  According to standard linguistic theory, the meaning of an utterance is the product of
  conventional semantic meaning and general pragmatic rules on language use. To investigate how
  cultural evolution of language plays out under this picture of the semantics-pragmatics
  interface, we present a game theoretic model of the competition between types of language
  users, each endowed with a selection of lexical representations and a particular pragmatic
  disposition to act on them. Our model traces two evolutionary forces and their interaction:
  (i) fitness-based pressure towards communicative efficiency and (ii) potential learning
  biases during the transfer of linguistic knowledge. We illustrate the model based on a case
  study on scalar implicatures. In this case study learning biases that favor simple semantic
  representations can foster the evolution of more sophisticated pragmatic reasoning types and
  so prevent the lexicalization of scalar implicatures.
\end{abstract}

@@

If this majority view is correct, the question arises how such a division of labor between
semantics and pragmatics could have evolved and why it would be so pervasive across natural
languages. Models of language evolution abound. There are simulation-based models studying the
evolution of language in populations of communicating agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of language evolution, mostly coming from game theory
\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
of this work has focused on explaining basic properties such as compositionality and
combinatoriality
\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax},
but little attention has been paid to the interaction between conventional meaning and
pragmatic use. What is more, many mathematical models explain evolved meaning as a regularity
in the overt behavior of agents, abstracting from complex interactions between semantic representations and pragmatic use. In contrast, we will here look at language users with a richer
cognitive make-up.

We spell out a model of the co-evolution of conventional meaning and pragmatic reasoning
types. The objects of replication and selection are pairs of lexical meanings and general types
of pragmatic behavior, which we represent using probabilistic models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. This
approach allows us to study the interaction between (i) evolutionary pressure towards
communicative efficiency and (ii) possible infidelity in the transmission of linguistic
knowledge, caused by factors such as inductive learning biases and sparse learning
data. Considering transmission of linguistic knowledge is important because neither semantic
meanings nor pragmatic usage patterns are directly observable. Instead, language learners have
to infer these unobservables from the observable behavior in which they result. We formalize
this process as a form of Bayesian inference. Our approach thereby contains a well-understood
model of iterated Bayesian learning \citep{griffiths+kalish:2007} as a special case, but
combines it with functional selection, here formalized as the most versatile dynamic from
evolutionary game theory; the replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this model.

@@

Pragmatic types define dispositions to produce and interpret messages given a lexicon. We
distinguish between two kinds of pragmatic types. {\em Literal interlocutors} produce and
interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
instead engage in mutual reasoning to inform their choices. Recent probabilistic models of
rational language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
level $0$, corresponds to literal language use, as in Equations~(\ref{h:level0}) and
(\ref{s:level0}). Pragmatic language users of level $n + 1$ act (approximately) rational with
respect to level-$n$ behavior of their interlocutors, as in Equations~(\ref{h:leveln}) and
(\ref{s:leveln}).
\begin{alignat}{3}
&H_{0}(s \mid m;L) &&\propto pr(s) L_{sm} \label{h:level0}\\
&S_{0}(m \mid s;L) &&\propto \exp(\lambda \; L_{sm}) \label{s:level0}\\
&H_{n+1}(s|m;L) &&\propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) &&\propto  \exp(\lambda \; H_{n}(s|m;L)) \label{s:leveln}
\end{alignat}

According to (\ref{h:level0}), a literal hearer's interpretation of a message depends on her
lexicon and her prior over states, $pr \in \Delta(S)$, which is here assumed flat, for
simplicity. Literal interpreters thereby choose an arbitrary true interpretation for each
message. Pragmatic hearers, defined in (\ref{h:leveln}), instead use Bayes' rule to weigh
interpretations based on a conjecture about speaker behavior. Speaker behavior is regulated by
a soft-max parameter $\lambda$, $\lambda \geq 0$ \citep{luce:1959,sutton+barto:1998}. As
$\lambda$ increases, choices approximate strict maximization of expected utilities. Expected
utility of a message $\messg$ in state $\state$ for a level $n+1$ speaker is here defined as
$H_{n}(s|m;L)$, the probability that the hearer will assign to or choose the correct
meaning. For literal speakers, utility only tracks truthfulness. Literal speakers choose any
true message with equal probability but may send false messages as well with a probability
dependent on $\lambda$.

@@

It is particularly important to note that pragmatic agents of level 1 using $\Llack$ associate $\msome$
preferentially with $\ssome$, in contrast to their literal counterparts of level 0. This
association is not perfect, and usually less strong than what agents with a lexicalized upper
bound in $\Lbound$ can achieve, even without pragmatic reasoning. Higher order 
reasoning beyond level $1$ leads to stronger associations of $\msome$ and $\ssome$ also for the
hearer. Still, the case study presented in Section~\ref{sec:si-case-study} will consider
speaker and hearer behavior at levels 0 and 1, as the latter are the simplest pragmatic
reasoning types which show a tendency to communicatively attuned pragmatic enrichment. Using
only level 1 reasoning types and possibly small $\lambda$ is therefore a conservative choice that works against the
fitness-based selection of pragmatic language use for a notion of fitness based on
communicative success, which is introduced next.

@@

In biological evolution, where types are expressed genetically, transmission infidelity comes
into the picture through infrequent and mostly random genetic mutations. However, an agent's
lexicon and pragmatic reasoning behavior is likely not inherited genetically. They need to be
learned from observation. Concretely, when agents of type $j$ want to adopt or imitate the
linguistic behavior of type $i$, they observe the overt linguistic behavior of type $i$ and
need to infer the covert type that most likely produced the visible behavior. Iterated learning
is a process in which languages are learned repeatedly from the observation of linguistic
behavior of agents who have themselves acquired the language from observation and inference. In
the simplest case there is a single teacher and a single learner. After sufficient training the
learner becomes a teacher and produces behavior that serves as input for a new learner. Due to
the pressure towards learnability it exerts, iterated learning alone generally leads to simpler
and more regular languages (see \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016} for
recent surveys).

@@

The model has three parameters: $\lambda$ regulates the degree to which pragmatic speakers
choose messages that maximize the chance of communicative success; $k$ is length of
observations for each language learner; $l$ regulates where the learners' inference behavior
lies on a spectrum from probability matching to choice of the most likely parent type.

@@

The model of the previous section formalizes the evolutionary competition between different
pairs of lexical representations and ways of using them. This section looks at a case study on scalar
implicatures. It engages in a following formal thought experiment to address the question: if a
population of language users could freely combine different lexica with different pragmatic
strategies, what are conditions under which the majority view of scalar implicatures could have
evolved?

Recall that the majority view is that scalar implicatures are systematic non-lexicalized pragmatic
enrichments. Scalar implicature triggers like \emph{some}, \emph{warm} or \emph{may} are
semantically weak expressions for which logically stronger expressions are salient, e.g.,
\emph{all}, \emph{hot} or \emph{must}. For instance, {\em some} is entailed by {\em all}. If
the sentence ``Chris owns all of Johnny Cash's albums'' is true, then ``Chris owns some of
Johnny Cash's albums'' is also true. However, while weaker expressions such as {\em some} are
truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what
their use is normally taken to convey. Instead, the use of a less informative expression when a
more informative one could have been used can license a defeasible inference that stronger
alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). In this way, ``Chris owns some
of Johnny Cash's albums'' is strengthened to convey that she owns {\em some but not all}
albums. According to the majority view, this is a pragmatic inference, not part of the
conventional meaning.

@@

To fill the model from Section~\ref{sec:model} with life, we need to specify the sets of
states, messages and lexica (Section~\ref{sec:stat-mess-conc}). Additionally, we want to
explore the effects of a learning bias in favor of simple lexical representations. One way of
motivating and formalizing such a bias is introduced below in
Section~\ref{sec:an-induct-learn}.

@@

Recall that types are a combination of a lexicon and a manner of language use. We analyze the
model's predictions in populations of types with one of the two behaviors introduced earlier;
literal or pragmatic. The former correspond to level $0$ reasoners and the latter to level
$1$. Accordingly, we consider a total of $432$ types. Six are variants of pragmatic language
users with $\Llack$-like lexica. We refer to these as \emph{target types}, because they
represent lexica and language use that conform to the majority view of scalar
implicatures. Twelve types are either literal or pragmatic speakers with lexica of the
$\Lbound$ kind. We refer to these as \emph{competitor types}, because they are expected to be
the target types' main contenders in evolutionary competition. Finally, note that while variants of a kind
lexicalize the same representations and use them in the same way when paired with identical linguistic behavior,
they map different states to different messages. More informally, they speak different languages
that lexicalize the same concepts. Consequently, more often than not types of a kind fail to communicate with each other.

@@

Recall that there are three parameters: soft-max parameter $\lambda$ affects how strongly
speakers favor messages that appear best from their subjective point of view; the
bottleneck size $k$ influences how faithfully learners tend to identify their teacher type; $l$
defines the learners' disposition towards choosing the most likely teacher type from the
posterior distribution. We expect that competitor types (types with lexica of the kind
$\Lbound$) have a fitness advantage over target types (pragmatic agents with lexica of the kind
$\Llack$), especially for very low levels of $\lambda$. Selection based on fitness alone may
therefore not lead to prevalence of target types in the population. On the other hand, lexica
of type $\Llack$ are simpler than those of type $\Lbound$ by the postulated measure from
above. This may make them more likely to be adopted by learners, especially when $k$ is low so
that different teacher types are relatively indistinguishable based on their behavior and when
$l$ is high. Still, lexica of the kind $\Lall$ are in turn more likely \emph{a priori} than
lexica of the kind $\Llack$. Simulation results will shed light on the question whether target
types can emerge, and for which parameter constellations.

To better understand the workings of a pressure towards communicative efficiency and a pressure
towards learnability, we look at the behavior of the replicator and mutator step first in
isolation, and then in combination. All simulation runs are initialized with an arbitrary
distribution over types, constituting a population's first generation. All reported results are
the outcome of 50 update steps. These outcomes correspond to developmental plateaus in which
change is, if not absent, then at least very slow. In other words, even if the resulting states
do not correspond to an eventual attracting state, they characterize long almost stationary
states in which the system remains for a very long time. As specified in
\S\ref{sec:learnability}, the mutation matrix $Q$ can be obtained by considering all possible
state-message sequences of length $k$. Given that this is intractable for large $k$, the sets
of data learners are exposed to are approximated by sampling $250$ $k$-length sequences from
each type's production probabilities.

@@

\begin{figure}[t]
\centering
\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig1-onlyr}
\caption{Stacked proportion of target types, in shades of green, and majority types, when other
  than target type, in $5$ independent populations after $50$ generations under only a
  pressure for communicative success. The right-most plots zoom in on only the proportion of target types.}
\label{fig:only-R}
\end{figure}

As expected, selection based on communicative success is sensitive to $\lambda$ as it
influences signaling behavior, which in turn determines communicative success. This is
showcased in Figure \ref{fig:only-R}, which depicts the proportion of target types in $5$
independent populations after $50$ replicator steps. The plot also shows the proportion of the
\emph{majority type}, i.e., the type with the highest proportion in the final population. With
low $\lambda$ many types have very similar behavior, so that evolutionary selection lacks grip
and becomes very slow. The result is a very long transition with near stagnancy in a rather
homogeneous population with many types. Conversely, higher $\lambda$ promotes more rational
linguistic behavior, widening the gap in expressivity between types and promoting more
homogeneous populations. As suggested by Figure \ref{fig:only-R}, the majority in most
populations is not one of the six pragmatic $\Llack$-style types. That is, a pressure only for
communicative success does not lead to a prevalence of target types under any $\lambda$-value. For
instance, with $\lambda = 20$ $1000$ independent populations only had $11$ cases in which
the target type was the majority type, corresponding to a mean proportion of $.003$ across
populations. By contrast, in $913$ cases the majority types had $\Lbound$ with close to an even
share between literal ($454$) and pragmatic types ($459$), corresponding to a mean proportion
of about $.48$ taken together.

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig2-onlym-pr}
\caption{Stacked proportion of target types, in shades of green, and majority types, when other than target type, in $5$ independent populations after $50$ generations under only a pressure for learnability ($\lambda = 20, k = 5$). The types' prior is shown in
  the right-most plot. The top-most groupings correspond to types, literal and pragmatic, with lexica of the kind of $\Llack$, $\Lbound$, and $\Lall$.}
\label{fig:only-M}
\end{figure}

@@

Pressure for communicative success or learnability are not sufficient on their own to have a
single target type dominate the population. When pressured for communicative success, the slight
communicative advantage of $\Lbound$ users leads to the their prevalence. When pressured for
learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more
complex alternatives such as $\Lbound$. However, learnability alone does not foment the
propagation of a single target type across the population.

@@

Figure \ref{fig:rmd} illustrates the combined effects of both pressures for a sample of
$\lambda$ and $l$ values. These results show that an inductive learning bias for simpler
semantics in tandem with functional pressure can lead to the selection of a single target
type, and so to a division of labor between semantics and pragmatics.
The proportion of a single majority target type increases with $\lambda$ and
$l$. Pressure for communicative success magnifies the effects of iterated learning and dampens
the proliferation of types of a kind that are equal in expressivity {\em and} learnability. A
pressure towards learnability favors the transmission of simpler semantics and thereby selects
for pragmatic language use.

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth,height=10cm,keepaspectratio]{./plots/fig4-3d-incumbents-difference} %alternatively without '-3d'
\caption{Mean difference between proportion of highest target type and highest other type in $1000$ independent populations after $50$ generations under both pressures ($k = 5$). \tb{Note that this plot interpolates data, but I strongly doubt there's a hidden peak/valley that is not accounted for. Another option would be to make a more grid-like surface without interpolation, but the main point is to have a plot from which the trends that value combinations give rise to are easy to read off, which I hope this one does. Compare also with the alternative 1D heatmap in the appendix.}}
\label{fig:diff}
\end{figure}

As for the effect of the sequence length $k$ not mentioned so far, it influences populations in
a predictable way: small values lead to more heterogeneous populations that reflect the
learner's prior more faithfully. This is due to the fact that the likelihood that a small
sequence was produced by any type is relatively uniform (modulo prior). By contrast, larger
values increasingly allow learners to differentiate types with different signaling behaviors.

@@

With respect to (i), \citet{kirby+etal:2015} propose a comparable model of the interaction
between a lexicon's expressivity and its learnability. One of its main differences to our proposal lies in 
the former force exerting pressure only in the production of learning data. The pressure towards communicative success 
within a population central to replication is absent. Put differently, here we considered a pressure for mutual understanding 
that may indirectly select for expressive types, i.e., those that can convey states unequivocally, whereas Kirby et al. only
consider the bearing that the latter ability has on the production of learnable data. We see two main reasons
for considering communicative success rather than just expressivity, and to only considering it in learning. First, learnability alone can 
lead to the acquisition of functionally defective languages, as showcased by $\Lall$ (cf. \citealt{kirby+etal:2008,silvey+etal:2014}, and \citealt{fay+etal:2013} for review of laboratory results). Second and more importantly, types may be equally expressive but their performance as means of information transfer crucially depends not only on themselves but on the population they find themselves in (cf. the competition of target types in Figure \ref{fig:rmd} and their lack of competition in Figure \ref{fig:only-M}). That is, we contend that having an expressive lexicon that generates learnable data does not  in itself capture a type's arguably central communicative function of transferring information to peers. Acquiring a maximally expressive type is of little use when it fails to be understood in its population.

@@

\newpage

\appendix

\section{notes}

>>>>

% \author{%\bf NAME1 and NAME2\\
%     ( -- draft \today --- )
% }

@@

\begin{abstract}
  According to standard linguistic theory, the meaning of an utterance is the product of
  conventional semantic meaning and general pragmatic rules on language use. To investigate how
  cultural evolution of language plays out under this picture of the semantics-pragmatics
  interface, we present a game theoretic model of the competition between types of language
  users, each endowed with a selection of lexical representations and a particular pragmatic
  disposition to act on them. Our model traces two evolutionary forces and their interaction:
  (i) pressure towards communicative efficiency and (ii) learning biases during the transfer of
  linguistic knowledge. We illustrate the model based on a case study on scalar
  implicatures. In this case study learning biases that favor simple semantic representations
  can foster the evolution of more sophisticated pragmatic reasoning types and so prevent the
  lexicalization of scalar implicatures.
\end{abstract}

@@

If this majority view is correct, the question arises how such a division of labor between
semantics and pragmatics could have evolved and why it would be so pervasive across natural
languages. Models of language evolution abound. There are simulation-based models studying
populations of communicating agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of language evolution, mostly coming from game theory
\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
of this work has focused on explaining basic properties such as compositionality and
combinatoriality
\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax},
but little attention has been paid to the interaction between conventional meaning and
pragmatic use. What is more, many mathematical models explain evolved meaning as a regularity
in the overt behavior of agents, abstracting from complex interactions between semantic
representations and pragmatic use. In contrast, we will here look at language users with a
richer cognitive make-up.

We spell out a model of the co-evolution of conventional meaning and pragmatic reasoning
types. The objects of replication and selection are pairs consisting of a set of lexical
meanings and a manner of pragmatic behavior. The latter are represented using probabilistic
models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. This
approach allows us to study the interaction between (i) pressure towards communicative
efficiency and (ii) infidelity in the transmission of linguistic knowledge, caused by factors
such as inductive learning biases and sparse learning data. Considering transmission of
linguistic knowledge is important because neither semantic meanings nor pragmatic usage
patterns are directly observable. Instead, language learners have to infer these unobservables
from the observable behavior in which they result. We formalize this process as a form of
Bayesian inference. Our approach thereby contains as a special case a well-understood model of
iterated Bayesian learning \citep{griffiths+kalish:2007}, but combines it with functional
selection, here formalized as the most versatile dynamic from evolutionary game theory, the
replicator dynamic \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces
this model.

@@

Pragmatic strategies define dispositions to produce and interpret messages given a lexicon. We
distinguish between two kinds of pragmatic strategies. {\em Literal interlocutors} produce and
interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
instead engage in mutual reasoning to inform their choices. Recent probabilistic models of
rational language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
level $0$, corresponds to literal language use, as in Equations~(\ref{h:level0}) and
(\ref{s:level0}). Pragmatic language users of level $n + 1$ act (approximately) rational with
respect to level-$n$ behavior of their interlocutors, as in Equations~(\ref{h:leveln}) and
(\ref{s:leveln}).
\begin{alignat}{3}
&H_{0}(s \mid m;L) &&\propto pr(s) L_{sm} \label{h:level0}\\
&S_{0}(m \mid s;L) &&\propto \exp(\lambda \; L_{sm}) \label{s:level0}\\
&H_{n+1}(s|m;L) &&\propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) &&\propto  \exp(\lambda \; H_{n}(s|m;L)) \label{s:leveln}
\end{alignat}

According to (\ref{h:level0}), a literal hearer's interpretation of a message depends on her
lexicon and her prior over states, $pr \in \Delta(S)$, which is here assumed flat for
simplicity. Literal interpreters thereby choose an arbitrary true interpretation for each
message. Pragmatic hearers, defined in (\ref{h:leveln}), instead use Bayes' rule to weigh
interpretations based on a conjecture about speaker behavior. Speaker behavior is regulated by
a soft-max parameter $\lambda \geq 0$ \citep{luce:1959,sutton+barto:1998}. As
$\lambda$ increases, choices approximate strict maximization of expected utilities. Expected
utility of a message $\messg$ in state $\state$ for a level $n+1$ speaker is here defined as
$H_{n}(s|m;L)$, the probability that the hearer will assign to or choose the correct
meaning. For literal speakers, utility only tracks truthfulness. Literal speakers choose any
true message with equal probability but may send false messages as well with a probability
dependent on $\lambda$.

@@

It is particularly important to note that pragmatic agents of level 1 using $\Llack$ associate
$\msome$ preferentially with $\ssome$, in contrast to their literal counterparts of level
0. This association is not perfect, and usually less strong than what agents with a lexicalized
upper bound in $\Lbound$ can achieve, even without pragmatic reasoning. Higher order reasoning
beyond level $1$ leads to stronger associations of $\msome$ and $\ssome$ also for the
hearer. Still, the case study presented in Section~\ref{sec:si-case-study} will consider
speaker and hearer behavior at levels 0 and 1, as the latter are the simplest pragmatic
reasoning types which show a tendency to communicatively attuned pragmatic enrichment. Using
only level 1 reasoning types and possibly small $\lambda$ is therefore a conservative choice
that works against the fitness-based selection of pragmatic language use for a notion of
fitness defined as communicative success, which is introduced next.

@@

In biological evolution, where types are expressed genetically, transmission infidelity comes
into the picture through infrequent and mostly random genetic mutations. However, an agent's
lexicon and pragmatic reasoning behavior is likely not inherited genetically. They need to be
learned from observation. Concretely, when agents of type $j$ want to adopt or imitate the
linguistic behavior of type $i$, they observe the overt linguistic behavior of type $i$ and
need to infer the covert type that most likely produced the observed behavior. Iterated learning
is a process in which languages are learned repeatedly from the observation of linguistic
behavior of agents who have themselves acquired the language from observation and inference. In
the simplest case there is a single teacher and a single learner. After sufficient training the
learner becomes a teacher and produces behavior that serves as input for a new learner. Due to
the pressure towards learnability it exerts, iterated learning alone generally leads to simpler
and more regular languages (see \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016} for
recent surveys).

@@

The model has three parameters: $\lambda$ regulates the degree to which pragmatic speakers
choose messages that appear optimal from the point of view of the agent's own utility measure
(which may be unrelated to the expected utility when communicating with a given population);
$k$ is length of observations for each language learner; $l$ regulates where the learners'
inference behavior lies on a spectrum from probability matching to choice of the most likely
parent type.

@@

The model of the previous section formalizes the evolutionary competition between different
sets of lexical representations and ways of using them. This section looks at a case study on
scalar implicatures. It engages in a formal thought experiment to address the
question: if a population of language users could freely combine different lexica with
different pragmatic strategies, what are conditions under which the majority view of scalar
implicatures could have evolved?

Recall that the majority view is that scalar implicatures are non-lexicalized pragmatic
enrichments. Scalar implicature triggers like \emph{some}, \emph{warm} or \emph{may} are
semantically weak expressions for which logically stronger expressions are salient, e.g.,
\emph{all}, \emph{hot} or \emph{must}. For instance, {\em some} is entailed by {\em all}. If
the sentence ``Chris owns all of Johnny Cash's albums'' is true, then ``Chris owns some of
Johnny Cash's albums'' is also true. However, while weaker expressions such as {\em some} are
truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what
their use is normally taken to convey. Instead, the use of a less informative expression when a
more informative one could have been used can license a defeasible inference that stronger
alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). In this way, ``Chris owns some
of Johnny Cash's albums'' is strengthened to convey that she owns {\em some but not all}
albums. According to the majority view, this is a pragmatic inference, not part of the
conventional meaning.

@@

To fill the model from Section~\ref{sec:model} with life, we need to specify the sets of
states, messages and lexica (Section~\ref{sec:stat-mess-conc}). Additionally, we want to
explore the effects of a learning bias in favor of simple lexical representations. One way of
motivating and formalizing such a bias is introduced in Section~\ref{sec:an-induct-learn}.

@@

Recall that types are a combination of a lexicon and a manner of language use. We analyze the
model's predictions in populations of types with one of the two behaviors introduced earlier;
literal or pragmatic. The former correspond to level $0$ reasoners and the latter to level
$1$. Accordingly, we consider a total of $432$ types. Six are variants of pragmatic language
users with $\Llack$-like lexica. We refer to these as \emph{target types}, because they
represent lexica and language use that conform to the majority view of scalar
implicatures. Twelve types are either literal or pragmatic speakers with lexica of the
$\Lbound$ kind. We refer to these as \emph{competitor types}, because they are expected to be
the target types' main contenders in evolutionary competition. Finally, note that while
different types may lexicalize the same representations, they may nevertheless map different
states to different overt messages. More informally, they speak different languages that
lexicalize the same concepts. Consequently, more often than not different variants of the same
kind may fail to communicate with each other.

@@

Recall that there are three parameters: soft-max parameter $\lambda$ affects how strongly
speakers favor messages that appear best from their subjective point of view; the
bottleneck size $k$ influences how faithfully learners can identify their teacher type; $l$
defines the learners' disposition towards choosing the most likely teacher type from the
posterior distribution. We expect that competitor types (types with lexica of the kind
$\Lbound$) have a fitness advantage over target types (pragmatic agents with lexica of the kind
$\Llack$), especially for very low levels of $\lambda$. Selection based on fitness alone may
therefore not lead to prevalence of target types in the population. On the other hand, lexica
of type $\Llack$ are simpler than those of type $\Lbound$ by the postulated measure from
above. This may make them more likely to be adopted by learners, especially when $k$ is low so
that different teacher types are relatively indistinguishable based on their behavior and when
$l$ is high. Still, lexica of the kind $\Lall$ are in turn even more likely \emph{a priori} than
lexica of the kind $\Llack$. Simulation results will shed light on the question whether target
types can emerge, and for which parameter constellations.

To better understand the workings of a pressure towards communicative efficiency and a pressure
towards learnability, we look at the behavior of the replicator and mutator step first in
isolation, and then in combination. All simulation runs are initialized with an arbitrary
distribution over types, constituting a population's first generation. All reported results are
the outcome of 50 update steps. These outcomes correspond to developmental plateaus in which
change is, if not absent, then at least very slow. In other words, even if the resulting states
do not correspond to an eventual attracting state, they characterize almost stationary states
in which the system remains for a very long time. As specified in \S\ref{sec:learnability}, the
mutation matrix $Q$ can be obtained by considering all possible state-message sequences of
length $k$. Given that this is intractable for large $k$, the sets of data which learners are
exposed to are approximated by sampling $250$ $k$-length sequences from each type's production
probabilities.

@@

\begin{figure}[t]
\centering
\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig1-onlyr}
\caption{Stacked proportion of target types, in shades of green, and majority types, when other
  than a target type, in $5$ independent populations after $50$ generations under only a
  pressure for communicative success. The right-most plots zoom in on only the proportion of
  target types.}
\label{fig:only-R}
\end{figure}

As expected, selection based on communicative success is sensitive to $\lambda$ since $\lambda$
influences signaling behavior, which in turn determines communicative success. This is
showcased in Figure \ref{fig:only-R}, which depicts the proportion of target types in $5$
independent populations after $50$ replicator steps. The plot also shows the proportion of the
\emph{majority type}, i.e., the type with the highest proportion in the final population. With
low $\lambda$ many types have very similar behavior, so that evolutionary selection lacks grip
and becomes very slow. The result is a very long transition with near stagnancy in a rather
homogeneous population with many types. Conversely, higher $\lambda$ promotes more
less-stochastic pragmatic behavior, widening the gap in expressivity between types and
promoting more homogeneous populations. As suggested by Figure \ref{fig:only-R}, the majority
in most populations is not one of the six pragmatic $\Llack$-style types. That is, a pressure
only for communicative success does not lead to a prevalence of target types under any
$\lambda$-value. For instance, with $\lambda = 20$ $1000$ independent populations only had $11$
cases in which the target type was the majority type, corresponding to a mean proportion of
$.003$ across populations. By contrast, in $913$ cases the majority types had $\Lbound$ with
close to an even share between literal ($454$) and pragmatic types ($459$), corresponding to a
mean proportion of about $.48$ taken together.

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig2-onlym-pr}
\caption{Stacked proportion of target types, in shades of green, and majority types, when other
  than a target type, in $5$ independent populations after $50$ generations under only a
  pressure for learnability ($\lambda = 20, k = 5$). The types' prior is shown in the
  right-most plot. The top-most groupings correspond to types, literal and pragmatic, with
  lexica of the kind of $\Llack$, $\Lbound$, and $\Lall$.}
\label{fig:only-M}
\end{figure}

@@

Pressure for communicative success and learnability are not sufficient on their own to have a
single target type dominate the population. When pressured for communicative success, the slight
communicative advantage of $\Lbound$ users leads to their prevalence. When pressured for
learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more
complex alternatives such as $\Lbound$. However, learnability alone does not foment the
propagation of a single target type across the population.

@@

Figure \ref{fig:rmd} illustrates the combined effects of both pressures for a sample of
$\lambda$ and $l$ values. These results show that an inductive learning bias for simpler
semantics in tandem with functional pressure can lead to the selection of a single target
type, and so to a division of labor between semantics and pragmatics.
The proportion of a single majority target type increases with $\lambda$ and
$l$. Pressure for communicative success magnifies the effects of iterated learning and dampens
the proliferation of multiple types of a kind that are equal in expressivity {\em and} learnability. A
pressure towards learnability favors the transmission of simpler semantics and thereby selects
for pragmatic language use.

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth,height=10cm,keepaspectratio]{./plots/fig4-3d-incumbents-difference} %alternatively without '-3d'
\caption{Mean difference between proportion of highest target type and highest other type in
  $1000$ independent populations after $50$ generations under both pressures ($k = 5$).}
\label{fig:diff}
\end{figure}

Effects of manipulating the sequence length $k$ have not been addressed so far, but are rather
predictable: small values lead to more heterogeneous populations that reflect the learner's
prior more faithfully. This is due to the fact that the likelihood that a small sequence was
produced by any type is relatively uniform (modulo prior). By contrast, larger values
increasingly allow learners to differentiate types with different signaling behaviors.

@@

With respect to (i), \citet{kirby+etal:2015} propose a comparable model of the interaction
between a lexicon's expressivity and its learnability. A main difference is that here we
considered a pressure for mutual understanding within a generation that may indirectly select
for expressive types, i.e., those that can convey states unequivocally, whereas Kirby et
al. only consider the bearing that the latter ability has on the production of learnable
data. We see two main reasons for considering communicative success rather than just
expressivity, and to only considering it in learning. First, learnability alone can lead to the
acquisition of functionally defective languages, as showcased by $\Lall$
(cf. \citealt{kirby+etal:2008,silvey+etal:2014}, and \citealt{fay+etal:2013} for review of
laboratory results). Second and more importantly, types may be equally expressive but their
performance as means of information transfer crucially depends not only on themselves but on
the population they find themselves in (cf. the competition of target types in Figure
\ref{fig:rmd} and their lack of competition in Figure \ref{fig:only-M}). That is, we contend
that having an expressive lexicon that generates learnable data does not in itself capture a
type's arguably central communicative function of transferring information to peers. Acquiring
a maximally expressive type is of little use when it fails to be understood in its population.

@@

% \newpage

% \appendix

% \begin{figure}[h!]
% \centering
% \includegraphics[width=1\textwidth,height=7.5cm,keepaspectratio]{./plots/fig4-incumbents-difference}
% \caption{Difference between mean proportion of highest pragmatic $\Llack$ type and highest
%   other type in $1000$ independent populations after $50$ generations under both pressures ($k
%   = 5$).}
% \end{figure}
